Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-04-02 summary.tex
first complete revision after barker & latter

\section{Summary and discussion}\label{summary}
In this paper we have studied the  axisymmetric vertical shear
instability (VSI) in astrophysical disks that arise from the
height dependence of the equilibrium angular velocity profile
$\Omega=\Omega(r,z)$ in a baroclinic disk. We examined how
the linear instability depends on disk thermodynamics and applied our
results to assess the relevance of the VSI in realistic protoplanetary
disks.

We generalized previous vertically global linear 
calculations of the VSI \citep{nelson13,mcnally14}, which assumed isothermal
perturbations, to include an energy equation. Our effort may also be
regarded as an extension of \cite{urpin03}, which did include an
energy equation, but adopted the Boussinesq and vertically local
approximations. We considered vertically global and fully
compressible perturbations.

We included a source term in the energy equation which relaxes the disk
temperature to its initial value on a timescale
$t_c=\beta\Omega_k^{-1}$. This approach allows the thermodynamic
response of the disk to be controlled by varying the parameter
$\beta$, and has been employed in previous numerical simulations of
the VSI \citep{nelson13}. However, it is possible to 
relate  $\beta$ to physical thermal timescales expected in realistic disks.

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. For
analytic discussion only, we further simplified the linear problem by
ignoring the background radial disk structure. This is equivalent to adopting
the `vertically global shearing box' (VGSB) formalism developed by
\cite{mcnally14}. 
We find the VGSB framework is appropriate for analyzing the VSI because 
rapid thermal relaxation ($\beta\ll1$) is 
generally required for the instability.   However, the VGSB framework is not
appropriate in the adiabatic limit ($\beta \gg 1$).

We mostly focused on vertically isothermal disks ($\Gamma=1$) as 
this is the relevant case for protoplanetary disks that are passively irradiated by its central star 
\citep{chiang97}. It also permits a largely analytical discussion.
Our analysis is an extension of the linear models considered in
\cite{lubow93} by  adding vertical shear and (rapid) thermal
relaxation. We obtained explicit solutions describing the VSI with 
thermal relaxation in the thin disk limit.

The fundamental VSI mode is characterized by a symmetric vertical
velocity perturbation with zero nodes and $\delta
v_z\sim\mathrm{constant}$ near the mid-plane, and by an anti-symmetric
density perturbation with one node at $z=0$.  These have been referred
to as `corrugation modes' by \cite{nelson13}, and were found to
dominate their numerical simulations, as well as that of
\cite{stoll14}. Furthermore, when computing unstable modes for a
protoplanetary disk model with thermal relaxation based on realistic
disk properties, we find the fundamental mode to be the dominant
mode. The fundamental mode is
thus likely the most relevant VSI mode in protoplanetary disks.

We determined the critical thermal relaxation timescale 
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below which small-scale
perturbations may be regarded as isothermal and the VSI 
operates. We have therefore provided a quantitative measure for the `fast'
thermal relaxation requirement for the VSI in astrophysical disks.

We also solved the linear stability problem numerically using a
pseudo-spectral approach. For numerical calculations we relax all the
approximations made in our analysis, and accounted for
the background radial disk structure.  Our
numerical results are mostly consistent with our analytic 
expectations, including the dependence of $\beta_\mathrm{crit}$ on
disk parameters.  We find the fundamental mode is robust against
boundary conditions, and higher order modes are more rapidly
stabilized with increasing $\beta$.

We briefly studied vertically non-isothermal disks. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI does not require rapid thermal
relaxation.  When a stable stratification is introduced, the
rapid reduction in VSI growth rates by a finite thermal relaxation
timescale is qualitatively similar to that observed for vertically
isothermal disks. Based on numerical results, we conjecture that the
critical thermal relaxation timescale is given by 
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-\Gamma)$ in vertically 
non-isothermal disks.

%application to MMSN

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks. We considered the Minimum Mass Solar Nebula 
(MMSN) described in \cite{chiang10} and   
modeled the thermal relaxation timescale $\beta$ based on realistic
disk properties and dust opacity. This results in a $\beta$ function that
depends on both the perturbation lengthscale as well as the height 
from the midplane. We find the MMSN is unstable to the VSI 
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. Although $\beta$ is not vertically constant in the
MMSN as assumed in our analysis, we find that evaluating the condition
$\beta <\beta_\mathrm{crit}$ at the disk midplane gives a rough measure of the
radii at which the VSI operates in the MMSN. This simple criterion is
thus a useful first step to estimate the relevance of the VSI in
realistic astrophysical disks without resorting to non-linear
numerical simulations.

@@

\subsubsection{Convective overstability}
We have focused on parameter regimes relevant to the VSI,
namely $\khat\gtrsim 1$ and $\beta\ll 1$. This excludes the
`convective overstability' recently discussed by \cite{klahr14} and
\cite{lyra14} in vertically unstratified disk models.

The convective overstability corresponds to slowly growing epicyclic motions.
It is most effective for $\beta\sim 1$ and disturbances with $|k_x|\ll |k_z|$, where
$k_z$ is a vertical wavenumber. These conditions are mutually exclusive with the VSI. 
The maximum growth rate for convective overstability is given by 
\begin{align}
  \nu_\mathrm{conv,max} = -\frac{N^2_r}{4\Omega}\label{max_conv_gen}
\end{align}
\citep{lyra14}, where $N_r^2 = -(c_p\rho)^{-1}\p_rP\p_rS$ is the
square of the radial buoyancy frequency
(cf. Eq. \ref{nzsq_def}). Thus, growth requires $N_r^2<0$.

@@

For more general disk structures convective overstability is
possible in principle. The models adopted for our numerical study in
\S\ref{numerical} do satisfy $N_r^2<0$, but convective overstability
require perturbations with long radial lengthscales,
$\khat\ll 1$, for which radially global disk models would be more 
self-consistent.

\subsubsection{Improved thermal evolution} 
We assumed non-adiabatic effects
can be represented by a thermal relaxation timescale $t_c$. For 
application to protoplanetary disks we connected $t_c$ to the disk
properties. One can improve the realism further by generalizing the 
opacity model adopted here to be density (and hence height) dependent. 
%here to be density-dependent,  i.e. $\kappa_d =
%% \kappa_d(\rho, T)$. 
However, going beyond thermal relaxation requires a
proper treatment of radiative diffusion. The linear problem then
involves a Laplacian operator and additional boundary
conditions must be supplemented. We will consider these improvements in a
follow-up study. %additional heat source

@@

We have neglected viscosity. \cite{nelson13} find that an 
alpha viscosity $\lesssim O(10^{-4})$ is required for the VSI. 
It is of interest to quantify the effect of viscosity 
on the linear VSI. This would help identify, more accurately, the
appropriate regions in a protoplanetary disk that are sufficiently
laminar to develop the VSI.  This problem is, however, complicated by
the fact that proper inclusion of viscosity introduces a meridional
flow in the equilibrium state, as well as viscous heating
in the energy equation.

>>>>

\section{Summary and discussion}\label{summary}
In this paper we have studied the axisymmetric vertical shear 
instability (VSI) in astrophysical disks that arise from the 
height dependence of the equilibrium angular velocity profile 
$\Omega=\Omega(r,z)$ in a baroclinic disk. Vertical shear is generally
expected in astrophysical disks due to the radial variation in the
disk temperature, or more generally the entropy. However, 
astrophysical disks are also typically sub-adiabatically stratified. 
In this case, the VSI requires a short thermal timescale $t_c$ in
order to operate.

Physically, a short thermal timescale is needed to overcome the 
stabilizing influence of buoynancy forces in a stably stratified
disk. Because vertical shear is weak in thin disks, $t_c\Omega\ll 1$
is generally needed for the VSI to operate, otherwise buoyancy is
strongly stabilizing. Our primary goal was to quantify this
requirement, thereby provide a simple way to assess 
whether or not the VSI is relevant in realistic disks.

To do so, we generalized previous vertically global linear 
calculations of the VSI \citep{nelson13,mcnally14,barker15}, which
assumed isothermal perturbations, to include an energy equation with 
a source term in that relaxes the disk
temperature to its initial value on a timescale
$t_c$. This approach of `thermal relaxation' allows
the thermodynamic response of the disk to be controlled by varying
$t_c$, and has been employed in previous numerical simulations of the
VSI \citep{nelson13}.

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. 
We mostly focused on vertically isothermal disks as 
this is the relevant case for protoplanetary disks passively
irradiated by its central star \citep{chiang97}. It also permits a
largely analytical discussion. Our analysis is an extension of the
linear models considered in \cite{lubow93} by adding vertical shear
and (rapid) thermal relaxation. We obtained explicit solutions
describing the VSI with thermal relaxation in the thin disk limit.

For a vertically isothermal, sub-adiabatically stratified disk we
determined the thermal timescale   
\begin{align*}
  t_\mathrm{crit} = \frac{\epsilon |q|}{\gamma-1}\Omega^{-1}, 
\end{align*}
below which the VSI operates effectively. While $t_\mathrm{crit}$ was
derived for the lowest order VSI mode --- the `fundamental  
corrugation mode' --- we find from numerical calculations that
$t_\mathrm{crit}$ can be used to characterize the general
stabilization of the VSI by a finite thermal timescale. %That is, all
%VSI growth rates are reduced significantly as $\beta$ is increased
%from 
%zero to $\beta_\mathrm{crit}$.    
We have therefore provided a quantitative measure for the thermal
timescale requirement for the VSI in astrophysical 
disks. The thermal timescale must be shorter than the disk
dynamical timescale by $O(\epsilon)$, which is
quite short for thin disks ($\epsilon\ll 1$), because vertical shear is weak.

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks. We considered the Minimum Mass Solar Nebula 
(MMSN) described in \cite{chiang10} and   
modeled the thermal timescale $t_c$ based on realistic
disk properties and dust opacity. 
%This results in $t_c$ being a
% function of position and the perturbation lengthscale of interest. 
We find the MMSN is unstable to the VSI 
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. It is possible to have the VSI at smaller
radii, but only on very small radial lengthscales (signficantly shorter
than the disk scaleheight), and such perturbations may
be subject to viscous decay in a real disk.

Although $t_c$ is not vertically constant in the 
MMSN as assumed in our analysis, we find that evaluating the condition
$t_c< t_\mathrm{crit}$ at the disk midplane gives an indication of the
radii at which the VSI operates in the MMSN. This simple criterion is
thus a practical first step in estimating the relevance of the VSI in
realistic astrophysical disks without resorting to linear
calculations or non-linear numerical simulations.

@@

\subsubsection{Convective overstability}
We have focused on parameter regimes relevant to the VSI,
namely perturbations with small radial lengthscales 
($k_xH\gtrsim 1$) in disks with short thermal timescales
($t_c\Omega\ll 1$). This excludes the `convective overstability'
recently discussed by \cite{klahr14} and \cite{lyra14} in vertically
unstratified disk models.

The convective overstability corresponds to slowly growing epicyclic motions.
It is most effective for $t_c\Omega \sim 1$ and disturbances with
$|k_x|\ll |k_z|$, where $k_z$ is a vertical wavenumber. These
conditions are mutually exclusive with the VSI.  
The maximum growth rate for convective overstability is given by 
\begin{align}
  \nu_\mathrm{conv,max} = -\frac{N^2_r}{4\Omega}\label{max_conv_gen}
\end{align}
\citep{lyra14}, where $N_r^2 = -(c_p\rho)^{-1}\p_rP\p_rS$ is the
square of the radial buoyancy frequency
(cf. Eq. \ref{nzsq_def}). Thus, growth requires $N_r^2<0$.

@@

For more general disk structures convective overstability is
possible in principle. The models adopted for our numerical study in
\S\ref{numerical} do satisfy $N_r^2<0$, but convective overstability
require perturbations with long radial lengthscales,
$k_xH\ll 1$, for which radially global disk models would be more 
self-consistent.

\subsubsection{Improved thermal evolution} 
We assumed non-adiabatic effects can be represented by a thermal
relaxation timescale $t_c$. For  
application to protoplanetary disks we connected $t_c$ to the disk
properties such as surface density, temperature and opacity, making
$t_c$ depend on position and the perturbation 
lengthscales. One might improve the realism further by generalizing
the opacity model adopted here to be density (and hence height)
dependent.  
%here to be density-dependent,  i.e. $\kappa_d =
%% \kappa_d(\rho, T)$. 
However, going beyond thermal relaxation requires a
proper treatment of radiative diffusion. % The linear problem then
% involves a Laplacian operator and additional boundary
% conditions must be supplemented. 
We will consider these improvements in a
follow-up study. %additional heat source

@@

We have neglected viscosity. \cite{nelson13} find that an alpha
viscosity $\lesssim O(10^{-4})$ is required for the
VSI. \cite{barker15} found that inclusion of viscosity is needed for 
convergence when studying surface modes. %but not body modes? 
Viscosity may also affect the
thermal timescale requirement of the VSI, because viscous heating
needs to be self-consistently included in the energy equation. This
should be investigated in future works.

% It is of interest to quantify the effect of viscosity 
% on the linear VSI. This would help identify, more accurately, the
% appropriate regions in a protoplanetary disk that are sufficiently
% laminar to develop the VSI.  This problem is, however, complicated by
% the fact that proper inclusion of viscosity introduces a meridional
% flow in the equilibrium state, as well as viscous heating
% in the energy equation.
